---
title: "Interesting cases"
author: "Stefano Allesina"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE)
setwd("../code/")
source("glv.R")
setwd("../notes/")
```

## Case 1: pairwise are always present

In these simulations, we start with a GLV pairwise model with a feasible equilibrium, and we add HOIs such that the equilibrium is unchanged. What is interesting is to see whether the HOIs can change the stability of the equilibrium. In all case, **we initialize the system very close to equilibrium**.

### Stable pairwise, unstable pairwise + HOIs (modification)

In this particular instance, the pairwise equilibrium is stable, but becomes unstable when adding HOI

```{r}
set.seed(119)
p <- build_GLV_HOIs(4, mode = "stable", HOIs = "modification", zerosumBi = TRUE)
# dynamics pairwise
outpairs <- integrate_dynamics(p, model = "glv")
# dynamics pairwise + HOIs
outhois <- integrate_dynamics(p, model = "glv_hois")
# plot them together
show(plot_output(outpairs, outhois))
```

### Stable pairwise, stable pairwise + HOIs (modification)

```{r}
set.seed(5)
p <- build_GLV_HOIs(4, mode = "stable", HOIs = "modification", zerosumBi = TRUE)
# dynamics pairwise
outpairs <- integrate_dynamics(p, model = "glv")
# dynamics pairwise + HOIs
outhois <- integrate_dynamics(p, model = "glv_hois")
# plot them together
show(plot_output(outpairs, outhois))
```

## Unstable pairwise, stable pairwise + HOIs (modification)

```{r}
set.seed(13)
p <- build_GLV_HOIs(4, mode = "stable", HOIs = "modification", zerosumBi = TRUE)
# make barely unstable
eM <- max(Re(eigen(diag(p$xstar) %*% p$A, only.values = TRUE)$values))
diag(p$A) <- diag(p$A) -1.1 * eM * 1/p$xstar
# dynamics pairwise
outpairs <- integrate_dynamics(p, model = "glv", maxtime = 200)
# dynamics pairwise + HOIs
outhois <- integrate_dynamics(p, model = "glv_hois", maxtime = 200)
# plot them together
show(plot_output(outpairs, outhois) + geom_point(data = tibble(time = 0, abundance = p$xstar, variable = as.factor(1:4), model = "second")))
```

This goes toward another stable equilibrium

```{r}
set.seed(42)
p <- build_GLV_HOIs(4, mode = "stable", HOIs = "modification", zerosumBi = TRUE)
# make barely unstable
eM <- max(Re(eigen(diag(p$xstar) %*% p$A, only.values = TRUE)$values))
diag(p$A) <- diag(p$A) -1.1 * eM * 1/p$xstar
# dynamics pairwise
outpairs <- integrate_dynamics(p, model = "glv", maxtime = 200)
# dynamics pairwise + HOIs
outhois <- integrate_dynamics(p, model = "glv_hois", maxtime = 200)
# plot them together
show(plot_output(outpairs, outhois) + geom_point(data = tibble(time = 0, abundance = p$xstar, variable = as.factor(1:4), model = "first")))
```

### Neutral pairwise, unstable pairwise + HOIs (general)

```{r}
set.seed(1005)
p <- build_GLV_HOIs(4, mode = "neutral", HOIs = "general", zerosumBi = TRUE)
# dynamics pairwise
outpairs <- integrate_dynamics(p, model = "glv")
# dynamics pairwise + HOIs
outhois <- integrate_dynamics(p, model = "glv_hois")
# plot them together
show(plot_output(outpairs, outhois))
```

### Neutral pairwise, stable pairwise + HOIs (general)

```{r}
set.seed(122)
p <- build_GLV_HOIs(4, mode = "neutral", HOIs = "general", zerosumBi = TRUE)
# dynamics pairwise
outpairs <- integrate_dynamics(p, model = "glv", maxtime = 400)
# dynamics pairwise + HOIs
outhois <- integrate_dynamics(p, model = "glv_hois", maxtime = 400)
# plot them together
show(plot_output(outhois,outpairs))
```




